Regularised parameter selection in the non-stationary Hawkes process

### 2017-04-25 _paper_output/paper_sparse_hawkes.tex
much texing

@@

    \section{$\ell_1$-penalised non-stationary Hawkes branching
process}\label{ellux5f1-penalised-non-stationary-hawkes-branching-process}

@@

We are interested in the problem of simultaneous identification of the
parameters of a Hawkes point process and identifying a sparse driving
signal by means of $\ell_1$ penalised regression where, by assumption,
the driving noise has a different form. This is analaogous to a blind
deconvolution problem or a compressive sensing-style inverse problem,
but considerably more challenging; While the process model superfcially
resembles a linear system, there are few of the conveneient results that
would apply in a linear regression problem; and as a time series problem
it is difficult to choose our penalty parameters by standard
non-parametric means such as cross validation. The proposed method uses
a heuristically-motivated degrees-of-freedom regularisation parameter
selection, which we test through simulation. The proposed method works
well in practice.

@@

We define call a counting process $N$ indexed by $t$ a (TODO)
predictable Poisson point process if it has an associated
\emph{intensity} process $\lambda$, such that $\lambda$ is predictable
with respect to the filtration generated by $N$, i.e.

@@

The Hawkes process \cite{hawkes_point_1971} specifically, is the linear,
self-exciting point process; Concretely, given a "background" rate
$\mu:\mathbb{R}^+\rightarrow\mathbb{R}^+,$ and a non-negative influence
kernel with positive support
$\phi:\mathbb{R}^+\rightarrow\mathbb{R}^+\cap\{0\}$, the Hawkes process
has the following expression for its intensity:

\[
\begin{aligned}
\lambda(\mathcal{F(t)}^-_{(-\infty,t)}) &= \mu(t) + \phi_\theta * N\\
&= \mu(t)  + \int_{-\infty}^{\infty}\phi_\theta(t-s)dN(s)\\
&= \mu(t) + \int_{-\infty}^{t}\phi_\theta(t-s)dN(s)\\
&= \mu(t) + \sum_{t_i\lt t}\phi_\theta(t-t_i)
\end{aligned}
\]

In this problem, we are given a single realisation of the count process
$N$ on an interval $T=[t_0,t_N]$. We are interested in identifying, from
this data, two unknown functional parameters; both the background rate
function $\mu$ and the influence kernel $\phi$. Since this is a general
nonparametric problem we will use traditional regularisation techniques
to control the expolosion of parameters.

    We hope to recover the excitation kernel, which is unknown, and also a
background signal, by appropriate penalisation of the terms.

For the penalized log likelihood for parameter $\theta$ and penalty
$\pi$, I write

\[L_\pi(\mathbf t, \theta'):= L_\pi(\mathbf t, \theta')- \pi P(\theta')\]

$P$ here is a non-negative functional which penalizes certain values of
the parameter vector, and $\pi\geq 0$ is the penalty weight
hyperparameter.

As before, the estimate is defined as the maximizer:

\[\hat{\theta}_\pi(\mathbf t) = \operatorname{argmax}_\theta L_\pi(\mathbf t;\theta)\]

I follow this practice here, penalizing only the deviations of
$\omega(t)$ from the parametric estimate.

    We take $\mu$ to be an unknown sparse, non-random signal in a particular
sense. We will assume that, given a linear basis
$\{\psi_j \}_{1\leq j\leq M}$, that we may write the background rate as
\[\mu(t) = \mu_0 + \sum_{j=1}^M  \omega_j \psi_j(t).\] \emph{Sparsity}
here will imply that the coefficients $\omega_j$ will be mostly 0.

In practice, we choose
\[\psi_j(t) := \mathbb{I}_{[\tau_{j},\tau_{j+1})}(t)\] for some lattice
$\{\tau_j\},$ which will correspond to the assumption that "sparse"
triggers will have localised support in time; we are looking for short
"spikes".

Similarly, we will approximate $\phi$ in a different functional basis
${f}_{1\leq k \leq K}$, such that $\phi(s):= \sum_{k=1}^K f(s;k)$. This
choice is likely arbitrary, but many exist in the literature trading off
computational and statistical convenience, including Laguerre
polynomials \cite{ogata_linear_1982}, Exponential functions
\cite{schoenberg_consistent_2005,rambaldi_modeling_2015}, and step
functions \cite{eichler_graphical_2016} As a convenient choice, we
propose a basis of smooth unimodal non-negative functions with a
computationally convenient form and a single mode, the convenient
Maxwell distribution,
\[\phi_j(t):=\sqrt{\frac{2}{\pi}} \frac{t^2\exp\left(-\frac{t^2}{2\sigma_j^2}\right)}{\sigma_j^3}.\]

A motivation for these specific choices is a particular real-world
problem: modelling viral memes in social media. We suspect that online
popularity is governed by some mixture of organic social sharing,
modeled by the self-exciting $\phi$ component, mixed with intermittent
external boosts, such as when an internet meme is featured on an
external medium such as television.

    mean delay $\mu=2\tau \sqrt{\frac{2}{\pi}}$

mode $\sqrt{2}\tau$

scale $\sigma=\sqrt{\frac{\tau^2(3 \pi - 8)}{\pi}}$

We take fixed $M$-point lattice $\tilde{\tau}$ and $K$-fixed scales
$\tilde{\sigma}.$

The parameter vector to be estimated then is
$\theta:=(\mu, \tilde{\omega}, \tilde{\kappa}).$

@@

    The non-penalised log-likelihood function for a given realisation of the
data is given \cite{ozaki_etimating_1979}

\[L_\pi(\mathbf t;\theta):=-\int_0^T\lambda_\theta(t)dt + \int_0^T\log \lambda_\theta(t) dN_t - \pi\|\boldsymbol \omega_\theta(t)\|_1\]

Using the method of maximum likelihood we estimate our parameter vector
by numerical optimisation of
\[\hat{\theta}_\pi(\mathbf t) = \operatorname{argmax}_\theta L_\pi(\mathbf t;\theta).\]

    There is no closed-form solution for the optimum of the Hawkes model
likelihood, penalised or not. The derivatives of the unpenalised
likelihood are easy to calculate manually -{}-{}- see e.g.
\cite{ozaki_maximum_1979}. Here, however, we use automatic
differentiation to make more complex penalisation procedure convenient
to differentiate \cite{baydin_automatic_2014}. The optimisation proceeds
using a standard black-box based Truncated Newton algorithm; as more
specialised sparse regression algorithms such as Least-Angle Regression,
do not directly apply here.

    To remedy this, I use Sugiura's finite-sample-corrected version, the
\emph{AICc} \cite{sugiura_further_1978}.
\[{\mathrm {AIC}(L(X;\hat{\theta}))}=2d-2\ln(L)\]

\[{\mathrm {AICc}}:={\mathrm {AIC}}+{\frac {2d(d+1)}{N-d-1}}\]

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{paper_sparse_hawkes_files/paper_sparse_hawkes_14_0.pdf}
    \end{center}
    { \hspace*{\fill} \\}

@@

The traditional means of performing model selection in Hawkes-type
models e.g. \cite{ogata_estimation_1983}, uses \emph{degrees-of-freedom}
penalties \cite{efron_how_1886}. Specifically, the \emph{AIC}, the
Akaike Information Criterion \cite{akaike_likelihood_1981} is widely
popular.

The AIC formula, for a model $M$ fit to a given data set $X$, for
estimated parameter vector $\hat{\theta^M}$ with log likelihood $L$ and
degrees of freedom $d^M$

\[\mathrm{AIC}(X,M)=2d^M-2L^M(X,\hat{\theta}^M)\]

The degrees of freedom for many models taken to be equal to the
cardinality of the parameter vector $\theta$.

Given two ML fitted models, $(M_1,\hat{\theta}^M_1)$ and
$(M_2\,\hat{\theta}^M_2)$, the difference in their AIC values is an
estimate of the relative Kullback-Leibler divergence of the inferred
measures $\hat{\mu_1},\hat{\mu_2}$ from the unknown true distribution,
$\mu$ i.e.

However, the AIC does not immediately solve the problem of
regularizition parameter selection in maximum likelihood, where the
regularisation parameter may be a real value smoothly interpolating many
possible models. One can calculate an AIC for penalised models, but this
require onerous Hessian matrix inversions and becomes costly in high
dimensions, and in any case does not generalise readily to non-smooth
penalties. \cite{konishi_information_2008}

We proceed heuristically, based on the arguments from Lasso-type
linear-model regularisation. There, the number of non-zero coefficients
fit under a particular value of the regularization penalty $\pi$ in an
$\ell_1$ regression model provides an unbiassed estimate of the
effective degrees of freedom of the model \cite{zou_degrees_2007}

    This suggests an adjusted likelihood parameter of the following form:

\[
\widehat{\mathrm{AICc}}(X, \hat{\theta}_\pi) = {\frac {2Nd(\hat{\theta}_\pi) }{N-d(\hat{\theta}_\pi)-1}} - 2\ln(L(X;\hat{\theta}_\pi), X))
\]

where $d_\pi$ counts the number of non-zero coefficients estimated for
given regularization coefficient $\pi$.

    Extensive grid search for optimsal parameters is expensive, so we select
both regularisation parameters using optimal sampling by Gaussian
process regression of the combined objective function.

\cite{snoek_practical_2012} using the package \texttt{scikit-optimize}
\cite{andreh7_scikit-optimize_2017} .

@@

        AttributeError                            Traceback (most recent call last)

        /usr/lib/python3.4/pickle.py in \_getattribute(obj, name, allow\_qualname)
        271         try:
    --> 272             obj = getattr(obj, subpath)
        273         except AttributeError:

        AttributeError: 'module' object has no attribute 'objective'

    During handling of the above exception, another exception occurred:

        AttributeError                            Traceback (most recent call last)

        <ipython-input-12-3150368f60c1> in <module>()
          1 from skopt import load
          2 
    ----> 3 res\_gp = load(join(FIT\_DIR, 'additive\_cvge'))

        /home/nfs/z3420853/Virtualenvs/basic34/lib/python3.4/site-packages/skopt/utils.py in load(filename, **kwargs)
        122         Reconstructed OptimizeResult instance.
        123     """
    --> 124     return load\_(filename, **kwargs)

        /home/nfs/z3420853/Virtualenvs/basic34/lib/python3.4/site-packages/sklearn/externals/joblib/numpy\_pickle.py in load(filename, mmap\_mode)
        573                     return load\_compatibility(fobj)
        574 
    --> 575                 obj = \_unpickle(fobj, filename, mmap\_mode)
        576 
        577     return obj

        /home/nfs/z3420853/Virtualenvs/basic34/lib/python3.4/site-packages/sklearn/externals/joblib/numpy\_pickle.py in \_unpickle(fobj, filename, mmap\_mode)
        505     obj = None
        506     try:
    --> 507         obj = unpickler.load()
        508         if unpickler.compat\_mode:
        509             warnings.warn("The file '\%s' has been generated with a "

        /usr/lib/python3.4/pickle.py in load(self)
       1036                     raise EOFError
       1037                 assert isinstance(key, bytes\_types)
    -> 1038                 dispatch[key[0]](self)
       1039         except \_Stop as stopinst:
       1040             return stopinst.value

        /usr/lib/python3.4/pickle.py in load\_global(self)
       1323         module = self.readline()[:-1].decode("utf-8")
       1324         name = self.readline()[:-1].decode("utf-8")
    -> 1325         klass = self.find\_class(module, name)
       1326         self.append(klass)
       1327     dispatch[GLOBAL[0]] = load\_global

        /usr/lib/python3.4/pickle.py in find\_class(self, module, name)
       1375         \_\_import\_\_(module, level=0)
       1376         return \_getattribute(sys.modules[module], name,
    -> 1377                              allow\_qualname=self.proto >= 4)
       1378 
       1379     def load\_reduce(self):

        /usr/lib/python3.4/pickle.py in \_getattribute(obj, name, allow\_qualname)
        273         except AttributeError:
        274             raise AttributeError("Can't get attribute \{!r\} on \{!r\}"
    --> 275                                  .format(name, obj))
        276     return obj
        277

        AttributeError: Can't get attribute 'objective' on <module '\_\_main\_\_'>

@@

@@

\bibliographystyle{unsrt}
\bibliography{refs}

>>>>

\usepackage{apacite}

@@

    \section{\texorpdfstring{\(\ell_1\)-penalised non-stationary Hawkes
branching
process}{\textbackslash{}ell\_1-penalised non-stationary Hawkes branching process}}\label{ell_1-penalised-non-stationary-hawkes-branching-process}

@@

We are interested in the problem of simultaneous identification of the
parameters of a Hawkes point process and identifying a sparse driving
signal by means of \(\ell_1\) penalised regression where, by assumption,
the driving noise has a different form. This is analaogous to a blind
deconvolution problem, robust regression, or a compressive sensing-style
inverse problem, but considerably more challenging; While the process
model superfcially resembles a linear system, there are few of the
convenient results that would apply in a linear regression problem; and
as a time series problem it is difficult to choose our penalty
parameters by standard non-parametric means such as cross validation.
The proposed method uses a heuristically-motivated degrees-of-freedom
regularisation parameter selection, which we test through simulation.
The proposed method works well in practice.

@@

We are given a counting process \(N\) on \(\mathbb R\), and an interval
\([0,T]\). (TODO clarify fit interval versus data interval)

We write the time series either by its values
\(\mathbf{t}=\{t_1, t_2,\dots,t_N(T)\}\). We represent the time series
either by its values or its arrival times as notationally conveneient.
We call it a (TODO) predictable Poisson point process if it has an
associated \emph{intensity} process \(\lambda\), such that \(\lambda\)
is predictable with respect to the filtration generated by \(N\), i.e.
\cite{daley_introduction_2003}

@@

    The Hawkes process \cite{hawkes_point_1971} in particular, is the
linear, self-exciting point process; Concretely, given a "background"
rate \(\mu:\mathbb{R}^+\rightarrow\mathbb{R}^+,\) and a non-negative
influence kernel with positive support
\(\phi:\mathbb{R}\rightarrow\mathbb{R}^+\cap\{0\}\), the Hawkes process
has the following expression for its intensity:

\[
\begin{aligned}
\lambda(\mathcal{F(N(t))}^-_{(-\infty,t)}) &= \mu(t) + \phi * N\\
&= \mu(t)  + \int_{-\infty}^{\infty}\phi(t-s)dN(s)\\
&= \mu(t) + \int_{-\infty}^{t}\phi(t-s)dN(s)\\
&= \mu(t) + \sum_{t_i\lt t}\phi(t-t_i)
\end{aligned}
\]

For brevity, we write
\(\lambda(\mathcal{F(t)}^-_{(-\infty,N(t))})=\lambda(t)\).

    The log-likelihood function for a given realisation of the data is given
\cite{ozaki_etimating_1979}

\[L_\pi(N;\mu,\phi):=-\int_0^T\lambda(t;\mu,\phi)dt + \int_0^T\log \lambda(t;\mu,\phi) dN_t \]

Using the method of maximum likelihood we estimate our parameter vector
by numerical optimisation of
\[\hat{\theta}_\pi(N) = \operatorname{argmax}_{\mu,\phi} L_\pi(N;\theta).\]

    We take \(\mu\) to be an unknown sparse, non-random signal in a
particular sense. We will assume that, given a basis
\(\{\psi_j \}_{1\leq j\leq M}\), that we may write the background rate
as \[\mu(t) = \mu_0 + \sum_{j=1}^M  \omega_j \psi_j(t).\]
\emph{Sparsity} here will imply that the coefficients \(\omega_j\) will
be mostly 0.

In practice, we choose
\[\psi_j(t) := \mathbb{I}_{[\tau_{j},\tau_{j+1})}(t)\] for some lattice
\(\{\tau_j\},\) which will correspond to the assumption that "sparse"
triggers will have localised support in time; we are looking for short
"spikes".

Similarly, we will approximate \(\phi\) in a different functional basis
\(\{\bar{\psi}\}_{1\leq k \leq K}\), such that
\(\phi(t):= \sum_{k=1}^K \bar{\psi}(t;k)\). This basis choice is likely
arbitrary, but many exist in the literature trading off computational
and statistical convenience, including Laguerre polynomials
\cite{ogata_linear_1982}, Exponential functions
\cite{schoenberg_consistent_2005,rambaldi_modeling_2015}, and step
functions \cite{eichler_graphical_2016} As a convenient choice, we
propose a basis of smooth unimodal non-negative functions with a
computationally convenient form, the Maxwell density,
\[\bar{\psi}_j(t;\sigma):=\sqrt{\frac{2}{\pi}} \frac{t^2\exp\left(-\frac{t^2}{2\sigma^2}\right)}{\sigma^3}.\]
so that

\[\phi(t):= \sum_{j=1}^K \kappa_j\bar{\psi}(t;\sigma_j).\] Once again,
some vector \(\{\sigma_j\}_{j\leq K}\) of time-scale parameters are
chosen \emph{a priori}.

@@

    Since this is a general nonparametric problem we will use traditional
regularisation techniques to control estimates. We hope to recover the
unknown excitation kernel and a background intensity by appropriate
penalisation of the terms.

For the penalized log likelihood for penalty functional \(\pi\), we
write

\[L_\pi(N, \mu, \phi):= L(N; \mu, \phi)- \pi(\mu, \phi)\]

As in plain log-likelihood optimization, the estimate is defined as the
maximizer:

\[\hat{\theta}_\pi(N) = \operatorname{argmax}_\theta L_\pi(N;\theta)\]

    A motivation for these specific choices is a particular real-world
problem: modelling viral sharing and spreading in social media. We
suspect that online popularity is governed by some mixture of two
componenets:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  organic social sharing, modeled by the self-exciting \(\phi\)
  component, and
\item
  intermittent external boosts represented as variations in the the
  \(\mu\) component, such as when an internet meme is featured on an
  external medium such as television.
\end{enumerate}

Models such as the Hawkes process have frequently been applied in an
attempt to disambiguate these two
\cite{crane_robust_2008,mitchell_hawkes_2010,rizoiu_expecting_2017}.

    \(\ell_1\) type penalties are known for their ability to produce sparse
solutions, so we choose penalties of the form
\[\pi(\theta)=p\left\|\{\sigma_j\}\right\|_1+q\left\|\{\omega_j\}\right\|_1.\]
With the right choice of the parameters \(p\) and \(q\) we hope to
shrink unnecessary parameters to zero, identifying predictively
necessary ones, as in Lasso selection.

    There is no closed-form solution for the optimum of the Hawkes model
likelihood, penalised or not. The derivatives of the unpenalised
likelihood are easy to calculate manually -\/-\/-
\cite{ozaki_maximum_1979}. Here, however, we use automatic
differentiation to make more complex penalisation procedure convenient
to differentiate \cite{baydin_automatic_2014}. The optimisation proceeds
using a standard black-box based Truncated Newton algorithm; as more
specialised sparse regression algorithms such as Least-Angle Regression,
do not directly apply here.

@@

The traditional means of performing model selection in Hawkes-type
models e.g. \cite{ogata_estimation_1983}, uses \emph{degrees-of-freedom}
penalties \cite{efron_how_1886}. Specifically, Akaike's
\cite{akaike_likelihood_1981} Information Criterion, or \emph{AIC}, is
widely popular.

The AIC formula, for a given log-likelihood parameter estimate
\(\hat{\theta}(N)\), with 'degrees of freedom', \(d\) is given

\[\mathrm{AIC}(N,\hat{\theta}(N))=2d-2L(N,\hat{\theta}(N))\]

The degrees of freedom for parametric models is taken to be the number
of free parameters, equal to the cardinality of the parameter vector
\(\theta\).

    The AIC does not immediately solve the problem of regularized parameter
selection where the regularisation parameter may be a real value
smoothly interpolating many possible models. One can calculate an AIC
for penalised models, but this require onerous Hessian matrix inversions
and becomes costly in high dimensions, and in any case does not
generalise readily to non-smooth penalties.
\cite{konishi_information_2008}

We proceed heuristically, based on the degrees-of-freedom arguments from
Lasso-type linear-model regularisation. There, the number of non-zero
coefficients fit under a particular value of the regularization penalty
\(\pi\) in an \(\ell_1\) regression model provides an unbiassed estimate
of the effective degrees of freedom of the model
\cite{zou_degrees_2007}.

    This suggests, by analogy, an AIC estimate based on the number of
non-zero parameters for given regularization coefficients:

\[
\hat{d}\left(N,\hat{\theta},\pi\right)=\sum_j\mathbb 1\left\{\hat{\theta_j}(N,p,q)\neq 0\right\}
\]

    The suggested combined procedure is to combine, in an inner loop, the
fitting of model parameters for given values of \(p\) and \(q\), and
selecting the hyperparameters with the best overall AIC value.

    \subsection{Simulations}\label{simulations}

    We simulate data with parameters to match the stylized behaviour of the
social media time series of previous work
\cite{crane_robust_2008,mitchell_hawkes_2010,rizoiu_expecting_2017}, to
test the efficacy of the proposed method.

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{paper_sparse_hawkes_files/paper_sparse_hawkes_22_0.pdf}
    \end{center}
    { \hspace*{\fill} \\}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{paper_sparse_hawkes_files/paper_sparse_hawkes_22_1.pdf}
    \end{center}
    { \hspace*{\fill} \\}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{paper_sparse_hawkes_files/paper_sparse_hawkes_22_2.pdf}
    \end{center}
    { \hspace*{\fill} \\}

    Extensive grid search for optimsal parameters is expensive, so we select
both regularisation parameters using optimal sequential sampling by
Gaussian process regression of the combined objective function
\cite{snoek_practical_2012} using the software package
\href{https://github.com/scikit-optimize/scikit-optimize}{scikit-optimize}.

    \subsection{Conclusions}\label{conclusions}

    \subsection{Appendix and nonsense}\label{appendix-and-nonsense}

@@

        FileNotFoundError                         Traceback (most recent call last)

        <ipython-input-8-765ef5deb338> in <module>()
          1 from skopt import load
          2 
    ----> 3 res\_gp = load(join(FIT\_DIR, 'additive\_cvge.gp.bz2'))

        /usr/local/lib/python3.6/site-packages/skopt/utils.py in load(filename, **kwargs)
        122         Reconstructed OptimizeResult instance.
        123     """
    --> 124     return load\_(filename, **kwargs)

        /usr/local/lib/python3.6/site-packages/sklearn/externals/joblib/numpy\_pickle.py in load(filename, mmap\_mode)
        565             obj = \_unpickle(fobj)
        566     else:
    --> 567         with open(filename, 'rb') as f:
        568             with \_read\_fileobject(f, filename, mmap\_mode) as fobj:
        569                 if isinstance(fobj, \_basestring):

        FileNotFoundError: [Errno 2] No such file or directory: '/Users/dan/Datasets/branching\_process\_output/additive\_cvge.gp.bz2'

@@

    Given two ML fitted models, \((M_1,\hat{\theta}^M_1)\) and
\((M_2\,\hat{\theta}^M_2)\), the difference in their AIC values is an
estimate of the relative Kullback-Leibler divergence of the inferred
measures \(\hat{\mu_1},\hat{\mu_2}\) from the unknown true distribution,
\(\mu\) i.e.

    mean delay \(\mu=2\tau \sqrt{\frac{2}{\pi}}\)

mode \(\sqrt{2}\tau\)

scale \(\sigma=\sqrt{\frac{\tau^2(3 \pi - 8)}{\pi}}\)

We take fixed \(M\)-point lattice \(\tilde{\tau}\) and \(K\)-fixed
scales \(\tilde{\sigma}.\)

The parameter vector to be estimated then is
\(\theta:=(\mu, \tilde{\omega}, \tilde{\kappa}).\)

    The AIC is based on asymptotic arguments, and Sugiura's
finite-sample-corrected version \emph{AICc}, has lower bias for most
practical applications \cite{sugiura_further_1978}.

\[{\mathrm {AICc}}:={\mathrm {AIC}}+{\frac {2d(d+1)}{N_T-d-1}}\]

I follow this practice here, penalizing only the deviations of
\(\omega(t)\) from the parametric estimate.

\[L_\pi(\mathbf t;\mu,\phi):=-\int_0^T\lambda(t;\mu,\phi)dt + \int_0^T\log \lambda(t;\mu,\phi) dN_t - \pi\|\boldsymbol \omega_\theta(t)\|_1\]

@@

\bibliographystyle{apacite}
\bibliography{refs}
